---
title: "Movie Shot Scale"
author: "Tabitha Sutherland, Angus Lindsay, Keegan Renshaw"
date: "STAT394 — T2-2026"
output:
  slidy_presentation: default
  beamer_presentation:
    theme: Madrid
    colortheme: dolphin
    fontsize: 9pt
    latex_engine: xelatex
  ioslides_presentation: default
subtitle: null
---

```{r,include=F}
library(Rcpp)
library(emdbook)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)

dat <- read.csv("movie_shotscale.csv", stringsAsFactors = TRUE,
                fileEncoding = "latin1")

# Clean text columns to avoid UTF-8 errors
dat$Title    <- iconv(as.character(dat$Title),    from = "latin1", to = "UTF-8", sub = "?")
dat$Director <- iconv(as.character(dat$Director), from = "latin1", to = "UTF-8", sub = "?")
dat$Country  <- iconv(as.character(dat$Country),  from = "latin1", to = "UTF-8", sub = "?")
dat$Title    <- factor(dat$Title)
dat$Director <- factor(dat$Director)
dat$Country  <- factor(dat$Country)

num <- c("ASL","RA","POV","INS","BCU","CU","MCU","MS","MLS","LS","VLS")

```

## Outline

- Quantitative descriptive analysis
  - Data and missing values
  - Univariate summaries
  - Correlations
  - Summaries by country
- Surprising observations
  - Impossible ASL values
  - A genuine extreme: Rope
  - Extreme shot counts
- Graphical/qualitative analysis
  - Correlation matrix
  - Pairs
  - Boxplots
  - Principal components
  - Factor analysis
- Surprising observations
  - Mahalanobis distance
- Can we assume that the data follow a multivariate Normal distribution?
  - QQ-Norm plots
  - Anderson-Darling Test
  - Mahalanobis Distances
- Are there groups in the data?
- Summary

# Quantitative descriptive analysis

## The data

The dataset contains `r nrow(dat)` films and `r ncol(dat)` variables.

Variables: Title, Year, Director, Country, ASL, and ten shot counts (RA, POV, INS, BCU, CU, MCU, MS, MLS, LS, VLS).

Median year `r median(dat$Year, na.rm = TRUE)` (range `r min(dat$Year, na.rm = TRUE)` to `r max(dat$Year, na.rm = TRUE)`), so the sample is concentrated in the 1920s and 1930s.

## Missing values

```{r}
miss <- colSums(is.na(dat))
miss <- miss[miss > 0]
miss_tab <- data.frame(
  Variable = names(miss),
  Missing  = as.integer(miss)
)
kable(miss_tab, caption = "Variables with missing values")
```

Missingness is minimal: Year has 2 missing values and BCU has 1. Country has one blank level and one film with no country; Director has 16 blank entries.

## Sample composition

```{r}
tab <- sort(table(dat$Country), decreasing = TRUE)
kable(data.frame(Country = names(tab), Films = as.integer(tab)),
      caption = "Number of films per country")
```

The USA dominates (215 films, 55.4%), followed by Britain (63), Germany (55) and France (37); the remaining countries have 5 or fewer films.

## Univariate summaries

```{r}
desc <- data.frame(
  Variable = num,
  Mean     = round(sapply(dat[num], mean,   na.rm = TRUE), 2),
  Median   = round(sapply(dat[num], median, na.rm = TRUE), 2),
  SD       = round(sapply(dat[num], sd,     na.rm = TRUE), 2),
  IQR      = round(sapply(dat[num], IQR,    na.rm = TRUE), 2),
  Min      = round(sapply(dat[num], min,    na.rm = TRUE), 2),
  Max      = round(sapply(dat[num], max,    na.rm = TRUE), 2)
)
kable(desc, caption = "Univariate numerical summaries")
```

## Reading the univariate table

ASL is strongly right-skewed: mean 11.79 s, median 7.70 s, SD 50.59 s, max 1000 s, min 0 s.

All count variables are overdispersed (variance greater than the mean), so median and IQR are more informative than mean and SD.

Most frequent shot type on average: LS (114.9), then MLS (100.2), MS (82.3), CU (79.2), MCU (74.0).

Least frequent: RA (21.2), VLS (20.2), POV (5.9), INS (5.9).

## Overdispersion check

```{r}
vmr <- data.frame(
  Variable = num,
  Mean     = round(sapply(dat[num], mean, na.rm = TRUE), 1),
  Variance = round(sapply(dat[num], var,  na.rm = TRUE), 1),
  VMR      = round(sapply(dat[num], function(x) var(x, na.rm=TRUE)/mean(x, na.rm=TRUE)), 1)
)
kable(vmr, caption = "Variance-to-mean ratio (VMR)")
```

All VMR values are well above 1, confirming overdispersion.

## Correlations (Spearman)

```{r}
cor_mat <- round(cor(dat[num], use = "pairwise.complete.obs",
                     method = "spearman"), 2)
kable(cor_mat, caption = "Spearman correlation matrix")
```

## Reading the correlations

Strong positive associations:

- POV and INS: 0.74
- RA and CU: 0.68
- BCU and CU: 0.60

Strong negative associations:

- CU and MLS: −0.77
- CU and LS: −0.71
- MCU and LS: −0.63

ASL is negatively correlated with RA (−0.28) and CU (−0.38), and positively with MLS (0.46).

This suggests a clear contrast between films built on close-ups and reverse angles and films built on longer, more distant shots.

## Summaries by country

```{r}
by_country <- aggregate(dat[num],
                        by = list(Country = dat$Country),
                        FUN = mean, na.rm = TRUE)
big <- names(tab[tab >= 4])
by_country <- by_country[as.character(by_country$Country) %in% big,
                         c("Country","ASL","RA","CU","MLS","LS")]
by_country[, -1] <- round(by_country[, -1], 1)
kable(by_country, caption = "Mean values by country (n >= 4)")
```

Only the USA, Britain, Germany and France can be compared with any confidence.

# Surprising observations

## Impossible values: ASL = 0

```{r}
zeros <- dat[!is.na(dat$ASL) & dat$ASL == 0,
             c("Title","Year","Director","ASL")]
kable(zeros, caption = "Films with ASL = 0")
```

An average shot length of zero seconds is impossible. These are almost certainly data errors or missing values coded as 0, and should be set to NA before further analysis.

## A genuine extreme: Rope

```{r}
top <- dat[order(-dat$ASL), c("Title","Year","Director","ASL")][1, ]
kable(top, caption = "Film with the largest ASL")
```

Rope (Hitchcock, 1948) has ASL = 1000 s.

This is not an error. Hitchcock deliberately shot Rope in a very small number of long continuous takes. An average of about 16.7 minutes per shot is consistent with that technique.

Lesson: an outlier can be scientifically meaningful, not just noise.

## Extreme shot counts

```{r}
ext <- dat[dat$BCU > 150 | dat$CU > 250 | dat$LS > 300,
           c("Title","BCU","CU","LS")]
kable(ext, caption = "Extreme shot-count observations")
```

Two opposite stylistic poles:

Close-up dominated: Dallas – Who Done It? (CU = 322, LS = 21), Eastenders (CU = 268, LS = 12).

Long-shot dominated: Havsgamar (CU = 0, LS = 320), Karleken Segrar (CU = 0, LS = 301).

Note also that Coronation Street, Dallas and Eastenders are TV series, not films, which is a scope issue to discuss.

## Blank country

```{r}
blank <- dat[dat$Country == "",
             c("Title","Year","Director","Country")]
kable(blank, caption = "Film with a blank country")
```

One film has no country recorded and should be treated as Unknown or excluded.




```{r,include=F}
library(ggplot2)
library(ggcorrplot)
library(mvtnorm)
library(emdbook)
library(GGally)
library(ggbiplot)
library(nortest)
library(readr)
library(ggcorrplot)
library(ggthemes)
library(ggbiplot)
library(psych)
library(hexbin)
library(MASS)
library(reshape2)
library(ggExtra)

movies <- read.csv(file="movie_shotscale.csv")
movies$Title <- iconv(movies$Title, from = "", to = "UTF-8", sub = "")
movies$Director <- iconv(movies$Director, from = "", to = "UTF-8", sub = "")
movies <- data.frame(movies)
movies$Title <- as.factor(movies$Title)
movies$Director <- as.factor(movies$Director)
movies$Country <- as.factor(movies$Country)

movies<-movies[complete.cases(movies),]
count.ry<-table(movies$Country)
movies$Countries <- ifelse(
  movies$Country %in% names(count.ry[count.ry >= 4]),
  as.character(movies$Country),"Other")
movies$Countries <- factor(movies$Countries)
```



# Graphical/qualitative analysis

## Correlation matrix

```{r}
corr <- cor(movies[,-c((1:4),16)])
pmat <- cor_pmat(movies[,-c((1:4),16)])
ggcorrplot(
  corr,p.mat=pmat,
  insig="stars",method="circle",hc.order=TRUE,type="lower")
```

---

Average shot length does not depend on what shots are included.
Medium shots have weak correlations with other shots, versatile shots that can be used in movies with a lot of close ups or long shots.
Long shots have strong negative correlation with close up shots.
Reverse angles have strong positive correlation with close up shots, shows conversations and character reactions.

## Pairs

```{r}
ggpairs(movies,columns = c(6:8,10,12,14),
  ggplot2::aes(colour = Countries, alpha = .7),
  upper = list(continuous = wrap("cor", size = 2)))
```

---

Displays shot types by country not including "medium/big/very" variants of close up/long shot.
Movies from Italy have strong linear relationships between POV:reverse angle and inserted shot:reverse angle shots.
Movies from Germany have very few reverse angle shots.
Strong negative correlation between close up:long shot in all countries.

## Boxplots

```{r}
movies.melt <- melt(data=movies[,6:16],
id.vars = "Countries",
variable.name = "Shots")

ggplot(data=movies.melt, aes(x=Shots, y=value)) +
geom_boxplot(aes(col=Shots), notch = TRUE) +
facet_grid(~ Countries) +
ylab("No. Shots") +
theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
legend.position = "none")
```

---

USA has similar mean shots for most shot types, likely because there are a lot more USA movies in dataset so have much greater variety of movies.
All country groups except USA show a trend of higher shot counts as type moves from big close ups towards long shots, until very long shots which is consistently lower than big close up.

## Principal Components

```{r}
movies.PCA<-prcomp(movies[,5:15],center=TRUE,scale=TRUE)

ggbiplot(movies.PCA,obs.scale=1,var.scale=1,groups=movies$Countries,
         ellipse=TRUE,labels = movies$Title)
```

---

PC1: High MLS and VLS, low RA, CU, MCU
PC2: High INS and LS, low MCU and MS

Sweden is furthest removed from the other country groups, and along with France varies much more along PC2 than PC1
Groups Britain, Italy, USA, and Other vary more across PC1 than PC2. Germany varies similarly across both.


## Factor Analysis

```{r}
fa.movies <- factanal(movies[,5:15], factors = 4)
par(family = "serif", mar = c(2, 2, 2, 5))

fa.diagram(fa.movies$loadings,digits = 3,rsize = 0.5,cex = 0.7,e.size = 0.08)
```

---

Factor 1 contains 24% of variation, shows movies with high close up and reverse angle counts are associated with lower medium long shot counts.
ASL and POV have high uniqueness and no strong associations to any factors. Factors do not capture variation in ASL and POV

# Surprising observations

## Mahalanobis distance

```{r}
mu.hat <- colMeans(movies[,5:15])
Sigma.hat <- cov(movies[,5:15])

dM <- mahalanobis(movies[,5:15], center=mu.hat, cov=Sigma.hat)
upper.quantiles <- qchisq(c(.9, .95, .99), df=9)
movies$dM <- dM
movies$surprise <- cut(movies$dM,
breaks=c(0, upper.quantiles, Inf),
labels=c("Typical", "Somewhat", "Surprising", "Very"))
table(movies$surprise)
surprising<-movies[movies$surprise %in% c("Surprising", "Very"),]
head(surprising[order(surprising$dM, decreasing = TRUE),c("Title","Year","dM","surprise")],5)

ggpairs(movies, columns=5:15,
ggplot2::aes(col=surprise, alpha=.5),
upper = list(continuous = "density", combo = "box_no_facet")) +
ggplot2::scale_color_manual(values=c("lightgray", "green", "blue", "red")) +
ggplot2::theme(axis.text.x = element_text(angle=90, hjust=1))
```

---

Rope is by far the most surprising observation, directed for shots to be as long as could fit on film.
`r nrow(surprising)` surprising/very surprising observations.
Plot shows very surprising observations appear either at the extremes of number of shots, or with high densities around very few shots in interaction plots.




```{r, include=FALSE}
vars <- c("ASL", "RA", "POV", "INS", "BCU",
          "CU", "MCU", "MS", "MLS", "LS", "VLS")

movie <- na.omit(dat[, vars])

head(movie)
summary(movie)
```

## Can we assume the data follows a multivariate Normal distribution?

## QQ-Norm plots
The QQ-Norm plots for each variable to check for normality

For the data to follow a multivariate normal distribution, each individual component must be normally distributed. We check this by visualising a QQ-Norm against QQ-line plot to visualize whether the data follows the line. 

## Q-Q plots: Variables that show they are not normally distributed

```{r qq1, fig.width=8, fig.height=4.2}
vars1 <- c("INS", "BCU", "CU", "MLS", "LS", "VLS")

par(
  mfrow = c(2, 3),
  mar = c(1.8, 1.8, 1.5, 0.5),
  oma = c(0.5, 0.5, 0, 0),
  mgp = c(1.1, 0.3, 0),
  cex.axis = 0.55,
  cex.main = 0.7
)

for (v in vars1) {
  qqnorm(
    movie[[v]],
    main = v,
    xlab = "",
    ylab = "",
    pch = 14,
    cex = 0.3
  )
  qqline(movie[[v]], col = "red", lwd = 1.5)
}


```

## QQ-Normality - The variables that don't follow a normal distribution

- Based on this, we see that the variables INS, BCU, CU, MLS, LS, 
  and VLS are not normally distributed. 
- ALS also has an extreme outlier, producing a right-skew: 
```{r qq2, fig.width=4, fig.height=4}
par(mfrow = c(1, 1))
qqnorm(movie[["ASL"]], main ="ASL")
qqline(movie[["ASL"]], col = "red", lwd = 1.5)
```




## Anderson darling test of normality

We then perform an Anderson-Darling test of normality with a null hypothesis that the data does follow a normal distribution and the alternative hypothesis that the data does not follow a normal distribution. 
According to the results:
- ASL, POV, INS, and VLS all show significant evidence against Normality, while RA, BCU, CU, MCU, MS, MLS, and LS do not provide evidence against normality. Due to the fact that multiple individuals fail a Normality check, we have evidence that the data are not multivariate normal. 

```{r}
library(goftest)
set.seed(12)

# perform an Anderson-darling test
ad_p <- sapply(movie, function(x) {
  m <- mean(x); s <- sqrt(mean((x - m)^2))   # MLE sd
  ad.test(x, null = "pnorm", mean = m, sd = s, estimated = TRUE)$p.value # checking if it follows a pnorm distribution
})
round(ad_p, 4)

```

## Mahalanobis distance check for Multivariate Normality

```{r mahalanobisResults, include=FALSE}
mu.hat = colMeans(movie)
Sigma.hat = (cov(movie))

dM2 = mahalanobis(movie, center=mu.hat, cov=Sigma.hat)

p=ncol(movie)
q = qchisq(c(.90, 0.95, 0.99), df=p)
obs1 = mean(dM2 > q[1])
obs2= mean(dM2 > q[2])
obs3 = mean(dM2 > q[3])

obs1
obs2
obs3
```
The observed proportions of large Mahalanobis distances are greater than those expected under a multivariate normal distribution. These inconsistencies becomes particularly pronounced at the 95% and 99% cut-offs. The results therefore suggest that the movie data do not follow a multivariate normal distribution closely, with more extreme multivariate observations than would be expected under normality.



## Based on the observed results, can we assume the data follow a multivariate normal distribution?

To conclude, we may assume that the data does not follow a multivariate normal distribution. Taken together, the marginal and multivariate diagnostics indicate that multivariate Normality is not a reasonable assumption for these data.



## Are there groups in these data?

## looking for groups
The seven shot-scale variables describe how shots are distributed from big close-ups to very long shots:
BCU, CU, MCU, MS, MLS, LS, and VLS

Because the total number of these coded shots is approximately 500 for most films, we converted them to proportions before comparing films

```{r groups-setup, include=FALSE}
library(cluster)
library(MASS)
scale_vars <- c("BCU", "CU", "MCU", "MS", "MLS", "LS", "VLS")
# Make sure all seven variables are retained
shot_counts <- movie[, scale_vars, drop = FALSE]
rows = as.integer(rownames(movie))
yr = dat$Year[rows]


# Convert counts to proportions
shot_prop <- movie[, scale_vars] / rowSums(movie[, scale_vars])

# Standardise
shot_scaled <- scale(shot_prop)

hc = hclust(dist(shot_scaled), method="ward.D2")
groups2 = cutree(hc, k=2)
groups3 = cutree(hc, k=3)
```

We focus on shot scale rather than all 11 variables because:
- the seven shot-scale variables directly describe the visual composition
- RA, POV, and INS contain a lot of the same zero values
- ASL contains an extreme value for Rope, which will affect the distance-based clustering strongly. 





## Checking for any distinct clusters
```{r shotScaleCompasion, fig.height=4, fig.width=4}
pca_groups <- prcomp(shot_scaled)

plot(
  pca_groups$x[,1],
  pca_groups$x[,2],
  col = groups2,
  pch = 19,
  cex = 0.6,
  xlab = "PC1",
  ylab = "PC2",
  main = "Shot-scale composition"
)

legend(
  "topright",
  legend = c("Group 1", "Group 2"),
  col = 1:2,
  pch = 19,
  cex = 0.8
)
```


## Discussion of shot-scale comparison
The Hierarchical clustering can divide the films into two groups but the PCA plot shows substantial overlap between them. 
This suggests that the films do not form two clearly seperated populations. Instead, the clustering algorithm is dividing a more continuous pattern into two parts. 


## Shot scale comparison through time
```{r groups-year, fig.width=6, fig.height=4}
plot(
  yr,
  pca_groups$x[,1],
  pch = 19,
  cex = 0.55,
  xlab = "Year",
  ylab = "PC1",
  main = "Shot-scale composition through time"
)

abline(
  lm(pca_groups$x[,1] ~ yr),
  col = "red",
  lwd = 2
)

pc1_year_cor <- cor(
  pca_groups$x[,1],
  yr,
  use = "complete.obs"
)
```


## Taking a look at the smallest cluster
```{r}
# Use the three-cluster solution
cluster_sizes <- table(groups3)

small_group <- as.integer(
  names(which.min(cluster_sizes))
)

early_group <- data.frame(
  Title = dat$Title[rows],
  Year = yr,
  CU = movie$CU,
  MLS = movie$MLS,
  LS = movie$LS,
  Group = groups3
)

early_group <- early_group[
  early_group$Group == small_group,
]

early_group <- early_group[
  order(early_group$Year),
]

early_small <- head(
  early_group[, c("Title", "Year", "CU", "MLS", "LS")],
  6
)

knitr::kable(
  early_small,
  caption = "Examples from the distinctive early subset",
  format = "html",
  table.attr = 'style="font-size:70%;"'
)
```


## Are there groups?
No clear groups
- The clustering algorithms can divide the films into groups, but the groups are not strongly separated.
- The clearest structure:
- 1. a strong continuous change in shot-scale composition through time
- 2.a small distinctive subset of films from approximately 1912-1916


## Summary

Quantitative descriptive analysis:

- ASL and most counts are right-skewed and overdispersed, so use median and IQR.
- Strong correlations: CU with MLS (−0.77), POV with INS (+0.74).
- USA dominates; only USA, BRI, GER, FRA allow reliable country comparisons.

Qualitative analysis:

- Length of shots (ASL) is independent of types of shots.
- Movies with a lot of close ups and reverse shots include few long shots.
- Similar trends in shot types hold across all countries excluding USA

Surprising observations:
- Four films with impossible ASL = 0, treat as missing.
- Rope (ASL = 1000) is a real, meaningful extreme.

Can we assume the data follow a multivariate Normal distribution?
- We have sufficient evidence to assume that the data does not follow a multivariate Normal distribution

Are there any groups in these data?
- There are no clear groups in the data


\vspace{0.5em}
\centering
